SIMD\slash SIMT-vectorized Sparse Chemical Kinetic Jacobian\slash Thermo-chemical Source Term Evaluation

### 2018-02-16 derivations.tex
add sparsity characterization

\section{Governing equations}
\subsection{State variables}
The state vector for this derivation consists of the temperature, a thermodynamic state parameter (pressure or volume) and the number of moles of all species except the last in the model:
\loadeq{state}
where $T$ is the temperature, $V$ and $P$ the volume and pressure respectively, and $n_j$ the number of moles of the $j$th species in the model (containing $\ns$ total species).

From the ideal gas law, the number of moles of the final species is determined as:
\begin{align}
n &= \frac{V P}{T \Ru} = \sum_{i=1}^{\ns}{n_i}\; , \nonumber \\
n_{\ns} &= \frac{V P}{T \Ru} - \sum_{i=1}^{\ns - 1}{n_i}
\end{align}

\subsection{Thermo-chemical source terms}
The evolution of the thermo-chemical state variables of this system is described by a set of ordinary-differential equations:
\loadeq{dstate}

@@

\subsubsection{Third-body enhanced reactions}
For a third-body enhanced reaction~\cref{e:thd_mix} is differentiated with respect to temperature to give:
\begin{align}
 \label{e:dci_thd_dt}
 \lhs{dci_thd_mix}{\frac{\partial [X]_i }{\partial T }} &= \rhs{dci_thd_mix}{-\frac{[C] \alpha_{\ns,i}}{T}} & \cond{dci_thd_mix}{,}
\end{align}
and with respect to the state parameter:
\begin{subequations}
 \label{e:dci_thd_de}
 \begin{align}
 \lhs{dci_thd_mix}{\frac{\partial [X]_i }{\partial V }} &= \rhs{dci_thd_mix}{\frac{1}{V} \left([C] \alpha_{\ns,i} - [X]_{i}\right)} & \cond{dci_thd_mix}{\dconp}\\
 \lhs{dci_thd_mix}{\frac{\partial [X]_i }{\partial P }} &= \rhs{dci_thd_mix}{\frac{\alpha_{\ns,i}}{T \Ru}} & \cond{dci_thd_mix}{\dconv}
 \end{align}
\end{subequations}
and finally, with respect to the moles of species $j$:
\begin{align}
\label{e:dci_thd_dnj}
\lhs{dci_thd_mix}{\frac{\partial [X]_i }{\partial n_{j}}} &= \rhs{dci_thd_mix}{\frac{1}{V} \left(- \alpha_{\ns,i} + \alpha_{j,i}\right)} & \cond{dci_thd_mix}{.}
\end{align}

>>>>

\section{Governing equations}
\subsection{State variables}
\label{s:state}
The state vector for this derivation consists of the temperature, a thermodynamic state parameter (pressure or volume) and the number of moles of all species except the last in the model:
\loadeq{state}
where $T$ is the temperature, $V$ and $P$ the volume and pressure respectively, and $n_j$ the number of moles of the $j$th species in the model (containing $\ns$ total species).

From the ideal gas law, the number of moles of the final species is determined as:
\begin{align}
\label{e:last_species}
n &= \frac{V P}{T \Ru} = \sum_{i=1}^{\ns}{n_i}\; , \nonumber \\
n_{\ns} &= \frac{V P}{T \Ru} - \sum_{i=1}^{\ns - 1}{n_i}
\end{align}

\subsection{Thermo-chemical source terms}
\label{s:source}
The evolution of the thermo-chemical state variables of this system is described by a set of ordinary-differential equations:
\loadeq{dstate}

@@

\subsubsection{Third-body enhanced reactions}
\label{s:dci_thd}
For a third-body enhanced reaction~\cref{e:thd_mix} is differentiated with respect to temperature to give:
\begin{align}
 \label{e:dci_thd_dt}
 \lhs{dci_thd_mix}{\frac{\partial [X]_i }{\partial T }} &= \rhs{dci_thd_mix}{-\frac{[C] \alpha_{\ns,i}}{T}} & \cond{dci_thd_mix}{,}
\end{align}
and with respect to the state parameter:
\begin{subequations}
 \label{e:dci_thd_de}
 \begin{align}
 \lhs{dci_thd_mix}{\frac{\partial [X]_i }{\partial V }} &= \rhs{dci_thd_mix}{\frac{1}{V} \left([C] \alpha_{\ns,i} - [X]_{i}\right)} & \cond{dci_thd_mix}{\dconp}\\
 \lhs{dci_thd_mix}{\frac{\partial [X]_i }{\partial P }} &= \rhs{dci_thd_mix}{\frac{\alpha_{\ns,i}}{T \Ru}} & \cond{dci_thd_mix}{\dconv}
 \end{align}
\end{subequations}
and finally, with respect to the moles of species $j$:
\begin{align}
\label{e:dci_thd_dnj}
\lhs{dci_thd_mix}{\frac{\partial [X]_i }{\partial n_{j}}} &= \rhs{dci_thd_mix}{\frac{1}{V} \left(- \alpha_{\ns,i} + \alpha_{j,i}\right)} & \cond{dci_thd_mix}{.}
\end{align}

### 2018-02-16 paper.tex
add sparsity characterization

@@

\subsection{Thermo-chemical Source Terms and Jacobian}
This new version of \texttt{pyJac} is capable of evaluating the thermo-chemical source terms for using both the constant-pressure (\conp) or constant-volume (\conv) assumption\footnote{Note: in this context, the ``constant-pressure'' and ``constant-volume'' assumptions refer to evaluation within a reaction sub-step in the operator splitting scheme, rather than a general constant-pressure or constant-volume reactive-flow simulation.}
In this section, we will only outline a brief summary of the system evaluated by \texttt{pyJac}; the interested reader is referred to our complete derivations, attached as supplimental material to this document.

The thermo-chemical state vector derivation consists of the temperature, the non-constant thermodynamic state parameter (volume or pressure) and the number of moles of all species except the last in the chemical model which is typically taken to be the bath-gas (e.g., \ce{N2}):
\loadeq{state}
where $T$ is the temperature, $V$ and $P$ the volume and pressure respectively, and $n_j$ the number of moles of the $j$th species in the model (containing $\ns$ total species).
The evolution of the thermo-chemical state vector is described by a set of chemical kinetic ordinary differential equations:
\loadeq{dstate}

For both \conp~and \conv, the molar source terms are~\cite{TurnsStephenR2012Aitc}:
\loadeq{dmolar}
where $\dot{\omega}_k$ is the $kth$ species' overall production rate:
\begin{equation}
 \dot{\omega}_{k} = \sum_{i=1}^{\nr} \nu_{k, i} R_{i} c_{i}
\end{equation}
where $\nu_{k, i}$ is the net stoichiometic coefficient of species $k$ in reaction $i$, $R_{i}$ the net rate of progress of reaction $i$ (described in~\cref{deriv-s:rop_net} of the supplimentary material), and $c_{i}$ the pressure modification term i.e., for third-body (see~\cref{deriv-s:thdbody} of the supplimental), or fall-off\slash chemically-activated reactions (\cref{deriv-s:falloff} of the same).

@@

>>>>

% set figure path
\graphicspath{{./figures/}}

@@

\subsection{Thermo-chemical Source Terms and Jacobian}
This new version of \texttt{pyJac} is capable of evaluating the thermo-chemical source terms for using both the constant-pressure (\conp) or constant-volume (\conv) assumption\footnote{Note: in this context, the ``constant-pressure'' and ``constant-volume'' assumptions refer to evaluation within a reaction sub-step in the operator splitting scheme, rather than a general constant-pressure or constant-volume reactive-flow simulation.}
In this section, we will only outline a brief summary of the system evaluated by \texttt{pyJac}; the interested reader is referred to our complete derivations, attached as supplemental material to this document.

The thermo-chemical state vector derivation consists of the temperature, the non-constant thermodynamic state parameter (volume or pressure) and the number of moles of all species except the last in the chemical model which is typically taken to be the bath-gas (e.g., \ce{N2}):
\loadeq{state}
where $T$ is the temperature, $V$ and $P$ the volume and pressure respectively, and $n_j$ the number of moles of the $j$th species in the model (containing $\ns$ total species).

This state vector---inspired by~\cite{SCHWER2002270}---has a number of features that recommend its use.
First, the state vector results in highly sparse chemical kinetic Jacobians as will be detailed in~\cref{S:sparsity}.
Second, mass is explicitly conserved in this formulation owing to the fact that the moles of the final species is calculated from the ideal gas law, and the rate of change of the final species from conservation of mass (see \cref{deriv-s:state,deriv-s:source} of the supplemental material respectively); i.e., system is not over-constrained.
Finally, the chemical kinetic Jacobian changes relatively little between the \conp~and \conv~forms, making maintaining the code-base much simpler.
Although most current combustion codes do not use species moles as a state variable, to\slash from more common mass\slash mole-fractions and moles is straightforward, and once inside integration of an chemical kinetic initial value problem, the choice of variables no longer matter.

The evolution of the thermo-chemical state vector is described by a set of chemical kinetic ordinary differential equations:
\loadeq{dstate}

For both \conp~and \conv, the molar source terms are~\cite{TurnsStephenR2012Aitc}:
\loadeq{dmolar}
where $\dot{\omega}_k$ is the $kth$ species' overall production rate:
\begin{equation}
 \dot{\omega}_{k} = \sum_{i=1}^{\nr} \nu_{k, i} R_{i} c_{i}
\end{equation}
where $\nu_{k, i}$ is the net stoichiometic coefficient of species $k$ in reaction $i$, $R_{i}$ the net rate of progress of reaction $i$ (described in~\cref{deriv-s:rop_net} of the supplemental), and $c_{i}$ the pressure modification term i.e., for third-body (see~\cref{deriv-s:thdbody} of the supplimental), or fall-off\slash chemically-activated reactions (\cref{deriv-s:falloff} of the same).

@@

\subsection{Sparsity patterns}
\label{S:sparsity}

In general, the Jacobian matrices generated by \texttt{pyJac} are largely sparse with non-zero entries mainly corresponding to species that participate in the same reaction (or non-default efficiency third-body species in a reaction), and mostly dense rows \slash columns corresponding to the temperature and thermodynamic state parameter.
However, the explicit-mass conservation formulation of \texttt{pyJac} can introduce additional non-zero entries in two ways.
First, if the last-species in the model (i.e., the bath-gas) participates directly in a reaction, the derivative of the forward or reverse rate of progress with respect to any other species is non-zero (see~\cref{deriv-s:dri_dnj}).
Similarly, if the last-species has a third-body efficiency not equal to the default (unity), this will again force all the derivative of the pressure-modification term with respect to other species to be non-zero (see~\cref{deriv-s:dci_thd}).
Either case will result in an undesirable fully dense Jacobian row for all species that have a non-zero net stoichiometic coefficient in such a reaction.

However, \texttt{pyJac} allows the user to simply ignore these derivatives (via a command-line switch) to and avoid any negative adverse affects on the Jacobian sparsity.
The rational behind this choice is that many common implicit integration techniques (e.g., CVODE~\cite{cvode:2.8.2}) used in solution of chemical kinetic initial value problems are formulated around the assumption that the supplied Jacobian is approximate; this allows the Jacobian and its LU factorization to be reused for multiple internal integration time-steps, accelerating the solution process.
For such solvers, there is no need for the exact form of the Jacobian and the so-called ``appropriate'' form is preferable.
Hence, in this section we will detail the sparsity of both forms of the Jacobian for the chemical models tested.

In~\cref{F:jac_sparsity}, a graphical representation of the sparsity pattern for GRI-Mech 3.0 is displayed.
In particular we note that~\cref{F:jac_sparse_approx} is has several rows that are no longer fully-dense, as result of its approximate form -- these rows correspond to species directly interacting with \ce{N2}, largely in GRI-Mech 3.0's \ce{NOx} chemistry reactions.
Table~\cref{T:jac_sparsity} shows the density of the exact and approximate Jacobians for all mechanisms tested in this work.
The smallest model, \ce{H2}\slash\ce{CO} is very dense with \textasciitilde\SI{70}{\percent} of the Jacobian entries non-zero however this drops to just \textasciitilde\SI{50}{\percent} for GRI-Mech 3.0, and continues to decrease to \textasciitilde\SI{27}{\percent} for USC-Mech II and just \textasciitilde\SI{10}{\percent} for the isopentanol model.
The approximate Jacobian assumption drops the density of Jacobian by \textasciitilde\SIrange{3}{7}{\percent} for all models.

\begin{figure}[htb]
  \centering
  \begin{subfigure}{0.45\linewidth}
      \includegraphics[width=\textwidth]{ch4_sparsity_exact.pdf}
      \caption{The ``exact'' Jacobian.}
      \label{F:jac_sparse_exact}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.45\linewidth}
      \includegraphics[width=\textwidth]{ch4_sparsity_approximate.pdf}
      \caption{The ``approximate'' Jacobian.}
      \label{F:jac_sparse_approx}
  \end{subfigure}
  \caption{A graphical representation of the sparsity pattern of the chemical kinetic Jacobian generated by \texttt{pyJac} for GRI-Mech 3.0.
	   Black squares indicate a non-zero Jacobian entry, while white square correspond to an empty entry.
	   The numbers indicate the index of the entry in the state vector.}
  \label{F:jac_sparsity}
\end{figure}

\begin{table}[tbp]
\centering
\begin{tabular}{@{}l S[table-format=2.1] S[table-format=2.1]@{}}
\toprule
Model                 & \multicolumn{1}{c}{Exact Jacobian Density} & \multicolumn{1}{c}{Approximate Jacobian Density} \\
\midrule
\ce{H2}\slash \ce{CO} & \SI{71.4}{\percent} & \SI{68.4}{\percent} \\
GRI-Mech 3.0          & \SI{56.7}{\percent} & \SI{49.8}{\percent} \\
USC-Mech II           & \SI{28.2}{\percent} & \SI{26.4}{\percent} \\
\ce{iC5H11OH}         & \SI{11.5}{\percent} & \SI{7.98}{\percent} \\
\bottomrule
\end{tabular}
\caption{The density of the exact and approximate Jacobians generated by \texttt{pyJac} for the various models studied.}
\label{T:jac_sparsity}
\end{table}
